---
title: "LUC Reconstruction Analysis in Hector"
author: "Pablo Tomberlin"
date: "2025-07-24"
output: html_document
---
```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

Hello! Here's my land-use change (LUC) emissions analysis project using Hector, a simple earth system model. Follow along to see my work.

If you want to run this code yourself, you can clone the repostitory:
https://github.com/PabloTomberlin/LUC_Emissions_Analysis

Read in data and load necessary packages
```{r}
data <- read.csv("selected_GCP.csv")
data_FF <- read.csv("Fossil_Fuels.csv")
observations <- read.csv("GML_observations.csv")
library(tidyverse)
library(Metrics)
library(ggsci)
library(ggrepel)
data <- data %>% filter(!(is.na(year)))
```

Fossil fuel visualization of a lengthening time series (2014, 2019, 2024)
```{r}
data_FF$GCP <- as.factor(data_FF$GCP)
ggplot() +
  geom_line(data = data_FF,
            aes(year, value, color = GCP),
            linewidth = 0.8) +
  facet_wrap("GCP", scales = "fixed") +
  scale_color_discrete() +
  ggtitle("Fossil Fuel Emissions") +
  xlab("Year") +
  ylab("Billion tonnes C (GtC/yr)") +
  theme_bw() +
  theme(text = element_text(size = 12, family = "mono", face = "bold"))
```

Plot LUC emissions data for 3 key GCP releases (first year, historical extension, and most recent year)
```{r}
LUC_results <- data %>% filter(GCP %in% c(2007, 2015, 2024))
LUC_results$GCP <- as.factor(LUC_results$GCP)
ggplot() +
  geom_line(data = LUC_results,
            aes(year, value, group = GCP, color = GCP),
            linewidth = 0.8) +
  ggtitle("LUC Emissions") +
  xlab("Year") +
  ylab("Billion tonnes C (GtC/yr)") +
  theme_bw() +
  theme(text = element_text(size = 12, family = "mono", face = "bold")) +
  scale_color_tron()
```

Basic Hector run
```{r}
library(hector)
ini_file <- system.file("input/hector_ssp245.ini", package = "hector")
core <- newcore(ini_file)
run(core)
out_default <- fetchvars(core, 1850:2100, vars = c(CONCENTRATIONS_CO2(),
                                                   GLOBAL_TAS(),
                                                   VEG_C(),
                                                   SOIL_C(),
                                                   LUC_EMISSIONS()))
out_default$GCP <- "Hector Default"
```

LUC plot with Hector Default (SSP 2-45) as separate line
```{r}
LUC_default <- out_default %>%
  filter(variable == "luc_emissions")
ggplot() +
  geom_line(data = LUC_results,
            aes(year, value, group = GCP),
            color = "gray",
            linewidth = 0.8) +
  geom_label_repel(aes(label = "Hector Unmodified LUC"),
                   x = 2070,
                   y = 0.1,
                   color = "blue") +
  facet_wrap("variable", scales = "free") +
  geom_line(data = LUC_default,
            aes(year, value, group = GCP),
            color = "blue",
            linewidth = 0.8) +
  ggtitle("LUC Emissions") +
  xlab("Year") +
  ylab("Billion tonnes C (GtC/yr)") +
  theme_bw() +
  theme(text = element_text(size = 12, family = "mono", face = "bold"))
```

Create list and for loop to run all carbon budgets individually through Hector and extract projections for climate and carbon cycle variables and conditions
```{r}
all_gcp_years <- unique(data$GCP)
results <- list()
for(gcp_year in all_gcp_years){
  isolated_data <- data %>% filter(GCP == gcp_year)
  message("running ", gcp_year)
  message("I see ", nrow(isolated_data), " rows of data")
  setvar(core, isolated_data$year, LUC_EMISSIONS(), isolated_data$value,
         getunits(LUC_EMISSIONS()))
  reset(core)
  run(core)
  results[[as.character(gcp_year)]] <- fetchvars(core, 1850:2100, vars = c(CONCENTRATIONS_CO2(),
                                                                           GLOBAL_TAS(),
                                                                           VEG_C(),
                                                                           SOIL_C(),
                                                                           LUC_EMISSIONS(),
                                                                           OCEAN_C()))
  
}

```

Combine all data frames including the default model run into one large data frame
```{r}
results_gcp <- bind_rows(results, .id = "GCP")
results_df <- bind_rows(results_gcp, out_default)
head(results_df)
```

Plot global temperature projections
```{r}
projections_gcp <- results_df %>%
  filter(variable == "global_tas") %>%
  filter(GCP %in% c(2007, 2015, 2024))
projections_default <-  results_df %>%
  filter(variable == "global_tas") %>%
  filter(GCP == "Hector Default")
ggplot() +
  geom_line(data = projections_gcp,
            aes(year, value, color = GCP),
            linewidth = 0.8) +
  #facet_wrap("variable", scales = "free") +
  #geom_line(data = projections_default,
            #aes(year, value),
            #color = "pink",
            #linewidth = 0.8) +
  ggtitle("Global temperatures projections") +
  xlab("Year") +
  ylab("Degrees C") +
  theme_bw() +
  theme(text = element_text(size = 12, family = "mono", face = "bold")) +
  scale_color_tron()
```

Calculate projection uncertainty due to LUC emissions revisions and store in hector_uncertainty
```{r}
hector_uncertainty <- results_df %>%
  filter(year %in% c(2024, 2100)) %>%
  filter(variable %in% c("global_tas", "CO2_concentration")) %>%
  group_by(year, variable) %>%
  summarise(
    mean_value = mean(value),
    sd_value = sd(value),
    margin_error = 1.96 * (sd_value/sqrt(length(unique(results_df$GCP)))),
    spread = max(value) - min(value),
    .groups = "drop"
  )
print(hector_uncertainty)
```

Relative graph showing deviation each projection has from the average of all temperature projections and a black line to indicate where observations end
```{r}
results_temp <- results_df %>%
  filter(variable == "global_tas")
results_avg <- results_temp %>%
  group_by(year) %>%
  summarize(mean_value = mean(value))
full_df <- left_join(results_temp, results_avg, by = "year")
diff <- mutate(full_df, subtract = value - mean_value)
tail(diff)

ggplot() +
  geom_line(data = diff,
            aes(year, subtract, color = GCP),
            linewidth = 0.8) +
  geom_vline(xintercept = 2023, linetype = "dashed", color = "black", linewidth = 0.8) +
  geom_hline(yintercept = 0, linetype = "dashed", color = "yellow", linewidth = 0.8) +
  xlab("Year") +
  ylab("Degrees (C)") +
  ggtitle("Difference from average projected temperature") +
  theme_bw() +
  theme(text = element_text(size = 12, family = "sans", face = "bold"))
```

Ocean, soil, and vegetation carbon plots to determine the driver of long-lasting LUC effects
```{r}
soil_veg <- results_df %>%
  filter(variable %in% c("soil_c", "veg_c", "ocean_c"))
ggplot() +
  geom_line(data = soil_veg,
          aes(year, value, color = GCP),
          linewidth = 0.8) +
  facet_wrap("variable", scales = "free") +
  ggtitle("Soil and vegetation carbon projections") +
  xlab("Year") +
  ylab("Pg C") +
  theme_bw() +
  theme(text = element_text(size = 12, family = "mono", face = "bold"))
```

Matilda Analysis (parametric uncertainty) and calculating 95% confidence interval for 100 Matilda runs. This was used to compare how uncertainty of historical LUC emissions and uncertainty of parameters each contributes to overall CO2 projection uncertainty
```{r, message=FALSE}
library(matilda)
ini_file <- system.file("input/hector_ssp245.ini", package = "hector")
core <- newcore(ini_file)
param_sets <- generate_params(core, draws = 100)
m_results <- iterate_model(core, 
                         params = param_sets,
                         save_vars = c("CO2_concentration"), save_years = 1850:2100)
```

```{r}
head(m_results)
ggplot() +
  geom_line(data = m_results,
            aes(year, value, group = run_number)) +
  xlab("Year") +
  ylab("Value (ppmv CO2)") +
  ggtitle("Parametric Uncertainty") +
  theme_bw() +
  theme(text = element_text(size = 12, family = "mono", face = "bold"))

matilda_uncertainty <- m_results %>%
  filter(year %in% c(2024, 2100)) %>%
  group_by(year, variable) %>%
  summarise(
    mean_value = mean(value),
    sd_value = sd(value),
    margin_error = 1.96 * (sd_value/10),
    spread = max(value) - min(value),
    .groups = "drop"
  )
print(matilda_uncertainty)
```

## Reproducibility

```{r}
sessionInfo()
```

